---
title: 'Lab 2: Parallel Computing'
subtitle: "Introduction to Statistical Computing"
author: "Elena Conderana & Sergio Cuenca"
date: 
output:
  pdf_document: 
    toc: true
    number_sections: false
    latex_engine: pdflatex
---

```{r, include=FALSE}
# A hook to wrap output based on a linewidth chunk option
# From https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE, linewidth=79)
```

\newpage

This lab is to be done outside of class time. You may collaborate with one classmate, but you must identify yourself and his/her name above, in the author's field, and you must submit **your own** lab as this completed .Rmd file. 

Installing and loading packages 
===

In order to perform the exercise in this practice you should install and load the `doParallel` package. 

```{r}
library(doParallel)
```

Q1. Is parallelization worth it?
===

**1a.** First, let's check how many cores do you have in your computer. Create a new variable `no_cores` equal to the number of cores minus 1.

```{r q1a}
no_cores <- parallel::detectCores()-1
```

**1b.** Register the cores and prepare the clusters using the `registerDoParallel()` and `makeCluster()` function. This will allow to parallelize code in the following code chunks.

```{r q1b}
registerDoParallel(no_cores)
cl <- makeCluster(no_cores)
```

**1c.** Now, you have the following function which calculates the prime numbers from `1` to `n`. Use the `microbenchmark` package to check which is faster to calculate the prime numbers when n goes from 10 to 10000: `lapply`, a `for` loop, `parLapply` or a `foreach` loop. The `lapply` and the `for` loop have been written for you. Which function is faster?
```{r q1c}
library(microbenchmark)

getPrimeNumbers <- function(n) {  
   n <- as.integer(n)
   if(n > 1e6) stop("n too large")
   primes <- rep(TRUE, n)
   primes[1] <- FALSE
   last.prime <- 2L
   for(i in last.prime:floor(sqrt(n)))
   {
      primes[seq.int(2L*last.prime, n, last.prime)] <- FALSE
      last.prime <- last.prime + min(which(primes[(last.prime+1):n]))
   }
   which(primes)
}
n_vec <- 10:10000
lapplyPrimeNumbers <- function(n_vec) {
  result <- lapply(n_vec, getPrimeNumbers)  
}
forPrimeNumbers <- function(n_vec) {
  result <- list()
  for (n in n_vec) {
    result[[n]] <- getPrimeNumbers(n)
  }
}

parLapplyPrimeNumbers <- function(n_vec, cl) {
  clusterExport(cl, "getPrimeNumbers")
  result <- parLapply(cl, n_vec, getPrimeNumbers)
}

foreachPrimeNumbers <- function(n_vec) {
  result <- foreach(n = n_vec, .export = "getPrimeNumbers") %dopar% {
    getPrimeNumbers(n)
  }
}

bench_results <- microbenchmark(
  for_loop = forPrimeNumbers(n_vec),
  lapply = lapplyPrimeNumbers(n_vec),
  parLapply = parLapplyPrimeNumbers(n_vec, cl),
  foreach = foreachPrimeNumbers(n_vec),
  times = 10
)

print(bench_results)
boxplot(bench_results)

```

**1d** Remember to use stop the clusters in `cl` using the `stopCluster` function.
```{r q1d}
stopCluster(cl)
```

**Challenge 01.** Search around your computer for a sequential code that might be parallelized. Using the `doParallel` package, parallelize the code and calculate the speedup. If you cannot find any code to parallelize, use the following code:
```{r q1ch1}
x <- iris[which(iris[,5] != "setosa"), c(1,5)]
trials <- seq(1, 10000)
boot_fx <- function(trial) {
  ind <- sample(100, 100, replace=TRUE)
  result1 <- glm(x[ind,2]~x[ind,1], family=binomial(logit))
  r <- coefficients(result1)
  res <- rbind(data.frame(), r)
}

cl <- makeCluster(no_cores)
parLapplyBoot <- function(trial, cl) {
  clusterExport(cl, c("boot_fx","x"))
  result <- parLapply(cl, trial, boot_fx)
}

bench_results_boot <- microbenchmark(
  boot_fx = boot_fx(trials),
  parLapplyBoot = parLapplyBoot(trials, cl),
  times = 10
)

print(bench_results_boot)
boxplot(bench_results_boot)

stopCluster(cl)

```