---
title: 'Lab 3: Testing'
subtitle: "Introduction to Statistical Computing"
author: Elena Conderana & Sergio Cuenca
date: 24/02/2025
output:
  pdf_document: 
    toc: true
    number_sections: false
    latex_engine: pdflatex
---

```{r, include=FALSE}
# A hook to wrap output based on a linewidth chunk option
# From https://github.com/yihui/knitr-examples/blob/master/077-wrap-output.Rmd
library(knitr)
hook_output = knit_hooks$get('output')
knit_hooks$set(output = function(x, options) {
  # this hook is used only when the linewidth option is not NULL
  if (!is.null(n <- options$linewidth)) {
    x = knitr:::split_lines(x)
    # any lines wider than n should be wrapped
    if (any(nchar(x) > n)) x = strwrap(x, width = n)
    x = paste(x, collapse = '\n')
  }
  hook_output(x, options)
})
opts_chunk$set(cache=TRUE, autodep=TRUE, cache.comments=TRUE, linewidth=79)
```

\newpage

This lab is to be done outside of class time. You may collaborate with one classmate, but you must identify yourself and his/her name above, in the author's field, and you must submit **your own** lab as this completed .Rmd file. 

Installing and loading packages 
===

In order to perform the exercises in this practice you should install and load the `testthat` and the `covr` package. 

```{r}
install.packages("testthat")
install.packages("covr")
```

Load the installed packages.

```{r}
library(testthat)
library(covr)
```

Q1. Unitary tests
===

**1a.** Let's begin with an easy one. You have the following function to convert from Celsius to Fahrenheit. Just check that it works by using an unitary test using the `testthat` package. You may need to see the help for the function `?test_that`:

```{r q1a}
C_to_F <- function(C_temp){
    F_temp <- (C_temp * 9/5) + 32;
    return(F_temp);
}

test_that("Celsius to Fahrenheit correct", {
  expect_equal(C_to_F(0), 32)
  expect_equal(C_to_F(100), 212)
  expect_equal(C_to_F(-40), -40)
  expect_equal(C_to_F(37), 98.6, tolerance = 0.1)
})
```

**1b.** You just have learned to test a function. Review the concept with the function to convert from Fahrenheit to Celsius:

```{r q1b}
F_to_C <- function(F_temp){
    C_temp <- (F_temp - 32) * 5/9;
    return(C_temp);
}

test_that("Fahrenheit to Celsius correct", {
  expect_equal(F_to_C(32), 0)
  expect_equal(F_to_C(212), 100)
  expect_equal(F_to_C(-40), -40)
  expect_equal(F_to_C(98.6), 37, tolerance = 0.1)
})
```

Q2. Using a test file
===

**2a.** Test functions can be saved like any other R script file (with a .R extension), but with one caveat. A test R script should start with the prefix ‘test-’. A good way of doing this is to add the prefix to the name of file that stores the functions to be tested. In this case, your teacher have included the `C_to_F` and `F_to_C` function to the `temp_conversion.R` file. You shall include the tests created in the previous section to the file `test-temp_conversion.R`. After that, run the following command to run all the test files included in the working directory:

```{r q2a}
test_dir(".")
```

**2b.** In the test file provided there is a failing test. Correct it and check that the `test_dir(".")` command now runs smoothly.

```{r q2b}
test_that("Fahrenheit to Celsius correct", {
  temp_F <- F_to_C(50)
  expect_equal(temp_F, 10)
})
```

Q3. Complicate things
===

**3a. ** Now that you have the setup done, create a file called `test-my_awesome_function.R` with tests for the function `my_awesome_function()` included in `my_awesome_function.R` so that they:
* Try the function with erroneous data type, for example, using a string instead of a number.
* Try the function with vectorized inputs.
* Check that the results are numeric.

After that, modify the function in order than none of this scenarios produce an error in the tests. You might need to include `stop()` statements in failing cases.
Try the tests using the `test_file` function:

```{r q3a}
test_file("test-my_awesome_function.R")
```

Q4. Code coverage
===

**4a. ** Now that you know how to modify tests, let's obtain the code coverage for the `number_utility.R` file. In order to do that, we will use the `covr` package to obtain a code coverage report of this file using the test file `test-number_utility.R`. 

```{r q4a}
# Constants
number_utility <- "number_utility"
extension_r <- ".R"
extension_htm <- ".htm"

# Manage file names
code_file_name <- paste(number_utility, extension_r, sep="")
test_file_name <- paste("test-", code_file_name, sep="")
coverage_report_file_name <- paste("coverage_report_", number_utility, extension_htm, sep="")

# Run tests and generate Code Coverage Report
test_file(test_file_name)
res <- file_coverage(code_file_name, test_file_name)
print(res)
report(res, coverage_report_file_name)
```

The code coverage is a sad 47.37% using the current tests. Modify the test file in order to achieve at least a 65% of code coverage. However, a 100% can be achieved.
